---
title: "Analysis of COVID-era Apple Mobility Data"
author: "Marta Bogaczynska" 
co-author: "Jacklyn Cherry"
date: "9/27/2020"
output: html_document
bibliography: references.bib 
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE)
```
## Setup
```{r set-parameters}
# set the state to work with
state_to_analyze <- "New York"
# if a state has a space in it we need to sanitize it
# sanitizing the state name for output files
state_no_spaces <- gsub(state_to_analyze,
                        pattern = " ",
                        replacement = "_")
# set the name of the global data file to use
global_file_name_to_process <- "data/raw_data/applemobilitytrends-2020-10-03.csv"
#extract out the core name for output file naming
core_name <- tools::file_path_sans_ext(basename(global_file_name_to_process))
```
```{r load-packages-functions}
# load the packages needed
library("dplyr")
library("readr")
library("ggplot2")
library("tidyr")
library("lubridate")
library("devtools")
# load the functions needed
source("code/functions/subset_mobility_data_to_state.R")
source("code/functions/count_up_counties_cities_transportation_type.R")
source("code/functions/convert_mobility_wide_to_long.R")
```
## Background and Objectives (Project Description)
This rmarkdown file will use the packages `ggplot2` [@ggplot2] and `dplyr`
[@dplyr] to analyze the Apple mobility data from the time before the Covid-19
pandemic, through the initial shutdowns, and the continuing changes over the
course of the year.
The goal will be to create a subset to a particular state, tally up the number
of cities and counties in that state with mobility data, and then to make
several plots to visualize those data subsets.
The state that this document will analyze is `r state_to_analyze`.

## Data Processing (Subsetting Data)
# three functions to load in

```{r subset-state}
# subset out to the chosen state using the variables set above
state_subset <- subset_mobility_data_to_state(
  input_file_name = global_file_name_to_process,
  state_to_subset = state_to_analyze)
```
```{r make-subset-long}
# convert the subsetted data from wide to long format
long_subset <- convert_mobility_wide_to_long(
  input_file_name = paste0("output/subsetted_states_wide/",
                        core_name,
                        "_",
                        state_no_spaces,
                        ".csv"))
```

```{r tallied-subset}
# tally up the number of cities and counties in the state that have mobility
# data available
tallied_subset <- count_up_cities_counties(
  input_file_name = paste0("output/subsetted_states_wide/",
                        core_name,
                        "_",
                        state_no_spaces,
                        ".csv"))
```

## Figures
```{r plot-tallied}
# create a grouped barplot of the tallied data to include in the knitted
# document, and save that figure to the output/figures directory as a png
tallied_plot <- ggplot(data = tallied_subset,
                       aes(x = geo_type,
                           y = n,
                           fill = transportation_type)) +
  geom_col(position = position_dodge())  +
  labs(title = paste("Number of cities and/or counties in",
                     state_to_analyze,
                     "with mobility data"),
       x = "type of area",
       y = "number of areas")
ggsave(plot = tallied_plot,
       filename = paste0("output/figures/",
                         core_name,
                         state_no_spaces,
                         "_tallied_plot.png"))
tallied_plot
```
```{r plot-time-series}
# produce a line plot of relative driving mobility across the state with date
# on the x axis and mean relative mobility on the y axis
timeseries_plot <- long_subset %>%
  filter(transportation_type == "driving") %>%
  group_by(date) %>%
  summarize(mean_mobility = mean(count)) %>%
  ggplot(aes(x = lubridate::ymd(date),
             y = mean_mobility)) +
  geom_line() +
  labs(title = paste("Statewide mean relative mobility driving levels in",
                     state_to_analyze,
                     "during COVID"),
       x = "Date",
       Y = "mean relative mobility")
ggsave(plot = timeseries_plot,
       filename = paste0("output/figures/",
                         core_name,
                         "_",
                         state_no_spaces,
                         "_timeseries_plot.png"))
timeseries_plot
```
```{r box-plot}
# produce a box plot
mobility_distribution_plot <- long_subset %>%
  filter(geo_type == "county") %>%
  ggplot(aes(x = factor(lubridate::month(lubridate::ymd(date))),
         y = count)) +
  geom_boxplot(aes(color = transportation_type)) +
  labs(title = paste("Statewide mobility distribution by county and month in",
                     state_to_analyze,
                     "during COVID-19"),
       x = "Month",
       y = "Relative number of requests")
ggsave(plot = mobility_distribution_plot,
       filename = paste0("output/figures/",
                         core_name,
                         "_",
                         state_no_spaces,
                         "_mobility_distribution_plot.png"))
mobility_distribution_plot
```

## Session Info
```{r session-info}
devtools::session_info()
```
## Sources Cited